---
title: "Day 9: GMM and DPD"
author: "Robert W. Walker"
date: "July 30, 2026"
format:
   revealjs:
     theme: [custom.scss]
     scrollable: true
     logo: AGSMlogo.jpeg
     footer: "ESSSSDA26-2J: GMM/DPD"
     multiplex:
       url: 'https://multiplex-5adl.onrender.com'
     self-contained: true
     html-math-method: katex
     incremental: true
     slide-number: c/t
     transition: zoom
     preview-links: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=TRUE, comment=NA, prompt=FALSE,
                      fig.height=6, fig.width=6.5, fig.retina = 3,
                      dev = 'svg', dev.args = list(bg = "white"), eval=TRUE)
library(tidyverse)
library(kableExtra)
library(plm)

# Panel summary utility
panel.summary <- function(mydata, idim) {
  mydata <- mydata
  idim   <- as.vector(idim)
  identifier <- table(idim)
  pan.btw    <- matrix(NA, nrow = length(identifier), ncol = ncol(mydata))
  for (i in seq_along(identifier)) {
    id.val  <- as.numeric(names(identifier)[[i]])
    pan.btw[i, ] <- sapply(1:ncol(mydata),
      function(j) mean(mydata[idim == id.val, j], na.rm = TRUE))
  }
  pan.mean <- matrix(NA, nrow = nrow(mydata), ncol = ncol(mydata))
  for (i in seq_len(nrow(mydata))) {
    id.val <- idim[[i]]
    pan.mean[i, ] <- sapply(1:ncol(mydata),
      function(j) mean(mydata[idim == id.val, j], na.rm = TRUE))
  }
  res.with    <- mydata - pan.mean
  grand.mean  <- colMeans(mydata, na.rm = TRUE)
  gmmat       <- matrix(grand.mean, nrow = nrow(mydata), ncol = ncol(mydata), byrow = TRUE)
  tab <- matrix(NA, 8, ncol(mydata))
  colnames(tab) <- colnames(mydata)
  rownames(tab) <- c("Grand mean","S.D.","TSS","Between S.D.","BSS","Within S.D.","WSS","% Within")
  tab[1, ] <- grand.mean
  tab[2, ] <- apply(mydata, 2, sd, na.rm = TRUE)
  tab[3, ] <- colSums((mydata - gmmat)^2, na.rm = TRUE)
  tab[4, ] <- apply(pan.btw, 2, sd, na.rm = TRUE)
  tab[5, ] <- colSums((pan.mean - gmmat)^2, na.rm = TRUE)
  tab[6, ] <- apply(res.with, 2, sd, na.rm = TRUE)
  tab[7, ] <- colSums(res.with^2, na.rm = TRUE)
  tab[8, ] <- tab[7, ] / tab[3, ]
  list(sum.tab = round(tab, 4))
}
```


## Outline for Day 9

::: {.nonincremental}
1. **Motivation** — Why dynamic panel models?
2. **GMM Foundations** — Moments, weighting, and overidentification
3. **Anderson-Hsiao** — The baseline IV approach
4. **Arellano-Bond** — Difference GMM
5. **System GMM** — Arellano-Bover / Blundell-Bond
6. **Specification Testing** — Sargan, Hansen, AR tests, and diagnostics
7. **Bias-Corrected LSDV** — Kiviet and Bruno corrections
8. **Extensions** — Unbalanced panels, Panel VAR, and software
9. **Applied Example** — UK manufacturing employment
10. **Practical Guidance and Philosophy**
:::

## Why Dynamic Models?

Many social processes are inherently **persistent** — the past directly shapes the present

- **State dependence**: employment, conflict onset, regime type, trade agreements
- **Habit formation**: consumption, party identification, welfare participation
- **Path dependence**: institutional arrangements, development trajectories
- **Partial adjustment**: firms, governments, and voters adjust slowly toward targets

A static panel model that assumes $\text{Cov}(y_{i,t-1},\,\epsilon_{it}) = 0$ is **misspecified** for these processes

The autoregressive parameter $\gamma$ quantifies the degree of persistence and the speed of adjustment:

$$\text{Speed of adjustment} = 1 - \gamma; \quad \text{Long-run multiplier} = \frac{\beta}{1-\gamma}$$

Models that ignore dynamics misattribute persistence to contemporaneous covariates — a **serious source of omitted variable bias**

## The DPD Model

$$y_{it} = \gamma\, y_{i,t-1} + \mathbf{X}_{it}\boldsymbol{\beta} + \alpha_i + \epsilon_{it}$$

- $i = 1,\ldots,N$ units (large); $t = 1,\ldots,T$ periods (fixed and small)
- $\gamma$ — autoregressive parameter; $|\gamma| < 1$ required for stationarity
- $\mathbf{X}_{it}$ — time-varying covariates; may be strictly exogenous, predetermined, or endogenous
- $\alpha_i$ — unobserved **unit-specific effect**, constant over time; correlated with $y_{i,t-1}$ by construction
- $\epsilon_{it}$ — idiosyncratic error; assumed IID$(0,\sigma^2_\epsilon)$ across $i$ and $t$

**The core problem**: since $y_{i,t-1} = f(\alpha_i, \epsilon_{i,t-1}, \ldots)$, the lagged dependent variable is **correlated with both $\alpha_i$ and past errors**. Every standard panel estimator fails for a different reason.

The dataset is characterized by large $N$ and fixed (small) $T$ — the short panel setting

## Within and Between Variation

For any variable $x_{it}$, total variance decomposes as:

$$\underbrace{\text{Var}(x_{it})}_{\text{Total}} = \underbrace{\text{Var}(\bar{x}_i)}_{\text{Between}} + \underbrace{\text{Var}(x_{it}-\bar{x}_i)}_{\text{Within}}$$

This decomposition matters enormously for identification:

- **Between variation** — variation *across* units (exploited by pooled OLS, RE)
- **Within variation** — variation *within* units over time (exploited by FE/within estimator)

For the DPD model with large $N$ and fixed $T$:

- Even within variation is **contaminated**: $y_{i,t-1} - \bar{y}_i$ is correlated with $\epsilon_{it} - \bar{\epsilon}_i$
- The share of within vs. between variation in the covariates determines identification strength

```{r panel-var-decomp, echo=FALSE}
data(EmplUK)
EmplUK$n  <- log(EmplUK$emp);    EmplUK$w  <- log(EmplUK$wage)
EmplUK$k  <- log(EmplUK$capital);EmplUK$ys <- log(EmplUK$output)
ps <- panel.summary(EmplUK[, c("n","w","k","ys")], EmplUK$firm)
ps$sum.tab |>
  kbl(caption = "EmplUK: Variance decomposition of log-transformed variables") |>
  kable_styling(font_size = 14)
```

## Naïve Approaches and Why They Fail

**Pooled OLS**: regress $y_{it}$ on $y_{i,t-1}$ and $\mathbf{X}_{it}$

- $\alpha_i$ enters the composite error; $\text{Cov}(y_{i,t-1},\alpha_i) > 0 \Rightarrow$ **upward bias** in $\hat{\gamma}$

**Fixed Effects (Within estimator)**: subtract unit means to remove $\alpha_i$

- $\widetilde{y}_{i,t-1} = y_{i,t-1} - \bar{y}_i$ is correlated with $\widetilde{\epsilon}_{it} = \epsilon_{it} - \bar{\epsilon}_i$ because $\bar{\epsilon}_i$ contains $\epsilon_{i,t-1}$
- **Downward (Nickell) bias** — opposite direction from pooled OLS

**Random Effects**: assumes $\text{Cov}(\alpha_i, y_{i,t-1}) = 0$

- Violated by construction whenever the unit effect drives the autoregressive process

**First-differencing alone**: removes $\alpha_i$ but creates a new endogeneity problem (treated next)

A useful diagnostic: the FE and pooled OLS estimates **bracket** the true $\gamma$

## Nickell / Hurwicz Bias

The within-group estimator is **inconsistent** with fixed $T$, even as $N \to \infty$

**Nickell (1981)** derives the approximate bias:

$$\text{Bias}(\hat{\gamma}_{FE}) \approx -\frac{1+\gamma}{T-1}\cdot\left[1 + \frac{2\gamma(1-\gamma^T)}{T(1-\gamma)}\right]^{-1}$$

Key implications:

- Bias is of order $O(1/T)$ — negligible only for large $T$ (e.g. $T > 30$)
- **Negative sign**: FE under-estimates persistence — the opposite direction from OLS
- Magnitude grows as $\gamma \to 1$: near-unit-root processes suffer worst
- With $\gamma = 0.8$ and $T = 5$: the formula gives $-0.45 / 2.076 \approx -0.22$; FE reports $\approx 0.58$

**Practical upshot**: for the short panels typical in cross-national or firm-level data ($T = 5$–$15$), FE estimates of dynamic models are substantially distorted. GMM is needed.

# GMM Foundations

## The Method of Moments

**Classical method of moments**: set population moments equal to their sample analogs and solve

For OLS the identifying condition is $\mathbb{E}[\mathbf{X}'\boldsymbol{\epsilon}] = 0$:

$$\frac{1}{N}\mathbf{X}'\hat{\boldsymbol{\epsilon}} = \mathbf{0} \quad \Rightarrow \quad \hat{\boldsymbol{\beta}}_{OLS} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$$

With $K$ regressors and $K$ conditions this is **exactly identified** — one unique solution

**Instrumental variables** replace $\mathbf{X}$ with instruments $\mathbf{Z}$ satisfying:

$$\underbrace{\mathbb{E}[\mathbf{Z}'\boldsymbol{\epsilon}] = \mathbf{0}}_{\text{orthogonality/exclusion}} \qquad \underbrace{\text{Cov}(\mathbf{Z},\mathbf{X}) \neq \mathbf{0}}_{\text{relevance/first stage}}$$

With $L = K$ instruments: **2SLS** solves $\frac{1}{N}\mathbf{Z}'\hat{\boldsymbol{\epsilon}} = \mathbf{0}$ exactly

With $L > K$ instruments: **overidentified** — cannot satisfy all $L$ conditions simultaneously $\Rightarrow$ need GMM

In the dynamic panel context, the panel structure generates **many** valid moment conditions for free — far more than parameters to estimate

## From MM to GMM: Overidentification and the $J$-Test

With $L > K$ valid moment conditions, define the sample moment vector:

$$g_N(\boldsymbol{\theta}) = \frac{1}{N}\sum_{i=1}^{N}\mathbf{Z}_i'\boldsymbol{\epsilon}_i(\boldsymbol{\theta})$$

**GMM** finds $\hat{\boldsymbol{\theta}}$ that brings $g_N(\boldsymbol{\theta})$ as close to zero as possible:

$$\hat{\boldsymbol{\theta}}_{GMM} = \arg\min_{\boldsymbol{\theta}}\; g_N(\boldsymbol{\theta})'\,\mathbf{W}_N\, g_N(\boldsymbol{\theta})$$

where $\mathbf{W}_N$ is a positive-definite **weighting matrix** — choice of $\mathbf{W}_N$ determines efficiency

The $L - K$ overidentifying restrictions are **testable** via the **Sargan/Hansen $J$-statistic**:

$$J = N\cdot g_N(\hat{\boldsymbol{\theta}})'\hat{\mathbf{W}}_N\, g_N(\hat{\boldsymbol{\theta}}) \xrightarrow{d} \chi^2(L-K)$$

::: {.callout-warning}
With many instruments, the Hansen test has very low power and will often fail to reject even when instruments are invalid. A $p$-value $> 0.95$ signals **too many instruments**, not good identification.
:::

## One-Step, Two-Step, and Windmeijer (2005)

**One-step GMM**: use a fixed, non-data-dependent weighting matrix

- Common choice: $\mathbf{W}_N = (\mathbf{Z}'\mathbf{H}\mathbf{Z})^{-1}$ under homoskedasticity
- Robust under weaker assumptions; but not asymptotically efficient

**Two-step GMM**: estimate the optimal weight matrix from first-step residuals

$$\hat{\mathbf{W}}_N^{(2)} = \left(\frac{1}{N}\sum_i \mathbf{Z}_i'\hat{\boldsymbol{\epsilon}}_i^{(1)}\hat{\boldsymbol{\epsilon}}_i^{(1)'}\mathbf{Z}_i\right)^{-1}$$

Asymptotically efficient — but finite-sample standard errors are **severely downward biased**, especially with many instruments

**Windmeijer (2005)** derives an analytic bias correction to the two-step VCE accounting for the fact that $\hat{\mathbf{W}}^{(2)}$ is itself estimated

::: {.nonincremental}
| Step | R (`pgmm`) | Stata |
|------|-----------|-------|
| One-step | `model = "onestep"` | default |
| Two-step + Windmeijer | `model = "twosteps"`, `summary(..., robust=TRUE)` | `twostep vce(robust)` |
:::

**Always use Windmeijer-corrected SEs with two-step GMM.** The Sargan test is unavailable after `robust`; use the Hansen $J$-test instead.

## GMM-Type Instruments and Sequential Moment Conditions

The DPD instrument set combines two types of moment conditions:

**Standard (IV-type)**: from strictly exogenous variables $x_{it}$

$$\mathbb{E}[x_{is}\,\Delta\epsilon_{it}] = 0 \;\; \forall\, s,t \quad \Rightarrow \quad \text{instrument using } \Delta x_{it}$$

**GMM-type (sequential)**: from the lagged dependent variable (and predetermined regressors)

$$\mathbb{E}[y_{i,t-s}\,\Delta\epsilon_{it}] = 0 \;\; \text{for } s \geq 2 \quad \Rightarrow \quad \text{instrument using } y_{i,t-2}, y_{i,t-3},\ldots$$

These accumulate: at period $t$, all levels $y_{i1}, \ldots, y_{i,t-2}$ are valid instruments

In `pgmm` syntax:

- **Left of `|`**: covariates entering the model (with desired lag structure)
- **Right of `|`**: instruments; wrap GMM-type ones in `lag(var, from:to)`

The distinction between strictly exogenous, predetermined, and endogenous regressors controls how deep into the lag structure instruments are drawn — will reemerge shortly.

# The Anderson-Hsiao Estimator

## First Differencing: Removing $\alpha_i$ and the New Problem

Take first differences of the DPD model to eliminate the unit effect:

$$\Delta y_{it} = \gamma\,\Delta y_{i,t-1} + \Delta\mathbf{X}_{it}\boldsymbol{\beta} + \Delta\epsilon_{it}$$

where $\Delta y_{it} = y_{it} - y_{i,t-1}$ and $\Delta\alpha_i = 0$ — **the fixed effect is gone**

However a **new endogeneity problem** arises immediately:

$$\Delta y_{i,t-1} = y_{i,t-1} - y_{i,t-2}$$

Since $y_{i,t-1}$ is a function of $\epsilon_{i,t-1}$, and $\epsilon_{i,t-1}$ appears in $\Delta\epsilon_{it} = \epsilon_{it} - \epsilon_{i,t-1}$:

$$\text{Cov}\!\left(\Delta y_{i,t-1},\,\Delta\epsilon_{it}\right) = -\text{Var}(\epsilon_{i,t-1}) \neq 0$$

We need instruments for $\Delta y_{i,t-1}$ that satisfy: (1) **relevance** — correlated with $\Delta y_{i,t-1}$; (2) **exogeneity** — uncorrelated with $\Delta\epsilon_{it}$

Under IID $\epsilon_{it}$: $y_{i,t-2}$ and $\Delta y_{i,t-2}$ are both valid — they are correlated with $\Delta y_{i,t-1}$ but predetermined relative to $\Delta\epsilon_{it}$

## Anderson and Hsiao (1981)

**Two 2SLS strategies** using observations at $t \geq 3$:

**Strategy 1** — Instrument with $\Delta y_{i,t-2}$ (lagged differences):

$$\text{Cov}(\Delta y_{i,t-2},\,\Delta\epsilon_{it}) = \text{Cov}(y_{i,t-2} - y_{i,t-3},\;\epsilon_{it} - \epsilon_{i,t-1}) = 0 \;\; \checkmark$$

**Strategy 2** — Instrument with $y_{i,t-2}$ (lagged levels):

$$\text{Cov}(y_{i,t-2},\,\Delta\epsilon_{it}) = 0 \;\; \checkmark \quad \text{and typically stronger first stage}$$

**Wawro (2002) and Beck**: The AH estimator performs poorly in finite samples. Econometricians agree.

**Fundamental inefficiency**: AH uses exactly **one instrument per time period**, leaving all deeper lags on the table. There are $\frac{(T-2)(T-1)}{2}$ valid lagged-level instruments — AH uses only $T-2$ of them.

Arellano and Bond (1991) recast AH as a GMM problem, using **all valid** lagged levels simultaneously and weighting them optimally.

# Arellano-Bond: Difference GMM

## The Arellano-Bond Idea

**Holtz-Eakin, Newey, and Rosen (1988)** and **Arellano and Bond (1991)** recognized:

At each time period $t$, every level $y_{i,s}$ with $s \leq t-2$ is a valid instrument for $\Delta y_{i,t-1}$

The instrument set grows with $t$:

::: {.nonincremental}
| Period $t$ | Equation | Valid level instruments for $\Delta y_{i,t-1}$ |
|:---:|:---:|:---:|
| 3 | $\Delta y_{i3}$ | $y_{i1}$ |
| 4 | $\Delta y_{i4}$ | $y_{i1},\; y_{i2}$ |
| 5 | $\Delta y_{i5}$ | $y_{i1},\; y_{i2},\; y_{i3}$ |
| $T$ | $\Delta y_{iT}$ | $y_{i1},\ldots,\; y_{i,T-2}$ |
:::

Total GMM instruments from the DV — for a **1-lag** AR($p=1$) model:

$$\sum_{t=3}^{T}(t-2) = 1 + 2 + \cdots + (T-2) = \frac{(T-2)(T-1)}{2}$$

This grows **quadratically** in $T$; with $p$ lags the first valid instrument shifts to $t=p+2$, giving $\frac{(T-p-1)(T-p)}{2}$ (see the Proliferation slide)

All are used simultaneously via GMM; the weighting matrix handles the fact that they are not all equally informative

For strictly exogenous covariates, $\Delta X_{it}$ contributes additional standard instruments at each period

## Building the Instrument Matrix

The instrument matrix $\mathbf{Z}_i^{AB}$ for unit $i$ has a **block-diagonal triangular** structure:

$$\mathbf{Z}_i^{AB} = \begin{pmatrix}
y_{i1} & & & \\
 & y_{i1} & y_{i2} & \\
 & & y_{i1} & y_{i2} & y_{i3} \\
 & & & & \ddots
\end{pmatrix}$$

Each row corresponds to one first-differenced observation ($t = 3, 4, \ldots, T$)

The complete orthogonality condition:

$$\mathbb{E}\!\left[\mathbf{Z}_i^{AB\prime}\,\Delta\boldsymbol{\epsilon}_i\right] = \mathbf{0}$$

This is a system of $\frac{(T-2)(T-1)}{2}$ moment conditions (for a 1-lag model) for one parameter $\gamma$ — **heavily overidentified**, and GMM is needed to optimally combine them

**Crucial**: the moment conditions are derived from lagged levels, not from outside the panel. The panel structure itself provides identification for free.

## The $T=3$ Case: Wawro's Worked Example

True model: $y_{it} = \gamma y_{i,t-1} + \alpha_i + u_{it}$. Substituting $y_{i2}$ into the $t=3$ equation:

$$\Delta_{3,2}y_i = \underbrace{\gamma[\overbrace{\gamma y_{i1}+\alpha_i+u_{i2}}^{y_{i2}}]+\alpha_i+u_{i3}}_{y_{i3}} - \underbrace{\gamma y_{i1}+\alpha_i+u_{i2}}_{y_{i2}}$$

$$= (\gamma^2-\gamma)y_{i1} + \underbrace{\gamma\alpha_i}_{\delta_i} + \underbrace{u_{i3}+(\gamma-1)u_{i2}}_{u^*_{\Delta(3,2)}}$$

This is an estimating equation without an observable regressor — we need both differences

For $\Delta_{2,1}y_i$, treating the **initial condition as fixed**:

$$\Delta_{2,1}y_i = (\gamma-1)y_{i1} + \alpha_i + u_{i2}$$

Because $u_{i1}$ and $u_{i2}$ are independent under IID errors, $y_{i1}$ is a valid instrument for $\Delta_{2,1}y_i$. Under normality the differenced errors are normal and independent — the AH and AB conditions hold.

## Strict Exogeneity, Predetermined, and Endogenous Regressors

The treatment of $\mathbf{X}_{it}$ in the instrument matrix depends on the assumed feedback:

**Strictly exogenous**: $\mathbb{E}[x_{it}\,\epsilon_{is}] = 0\;\;\forall\, s,t$

- No feedback in any direction; all periods valid as instruments
- Enter the instrument matrix as differences: $\Delta x_{it}$

**Predetermined** (weakly exogenous): $\mathbb{E}[x_{it}\,\epsilon_{is}] \neq 0$ if $s < t$; $= 0$ if $s \geq t$

- Backward feedback allowed (past outcome affects $x$), no contemporaneous/future feedback
- Instrument with $x_{i,t-1},\; x_{i,t-2},\ldots$ — levels at $t-1$ and before

**Endogenous**: $\mathbb{E}[x_{it}\,\epsilon_{is}] \neq 0$ for $s \leq t$

- Contemporaneous correlation; instrument with $x_{i,t-2},\; x_{i,t-3},\ldots$

Misclassifying a predetermined regressor as strictly exogenous will **use invalid instruments** and produce inconsistent estimates. The Sargan/Hansen test may or may not catch this.

## Instrument Proliferation and Collapsing

For a model with $p$ lags of $y$, the first valid GMM instrument appears at $t = p+2$ (period $t$ must pre-date both endogenous terms). Total instruments from the DV:

$$\frac{(T-p-1)(T-p)}{2}$$

This is a **quadratic** function of $T$ — for $p=1$: $(T-2)(T-1)/2$; for $p=2$: $(T-3)(T-2)/2$

::: {.nonincremental}
| $T$ | $p=1$ lag: $(T-2)(T-1)/2$ | $p=2$ lags: $(T-3)(T-2)/2$ |
|:---:|:---:|:---:|
| 5 | 6 | 3 |
| 7 | 15 | 10 |
| 10 | 36 | 28 |
| 15 | 91 | 78 |
:::

The AB Table 4(a1) replication uses $p=2$, so the instrument counts in the right column apply. Note: additional predetermined covariates each contribute further columns with the same triangular structure — instrument count escalates rapidly.

**Consequences** (Roodman 2009): overfits the first stage; Hansen $p$-values cluster near 1.0; estimates biased toward FE. Rule of thumb: **instruments $\leq$ groups**.

**Remedy 1 — Limit lag depth**: `lag(log(emp), 2:4)` instead of `lag(log(emp), 2:99)`

**Remedy 2 — Collapse the instrument matrix** (Roodman): one column per lag depth rather than one column per lag-period combination; reduces instruments from $O(T^2)$ to $O(T)$

```stata
* In xtabond2 (Stata)
xtabond2 n L.n L(0/1).w, gmm(L.n, lag(2 4) collapse) iv(L(0/1).w) twostep robust
```

Report both collapsed and uncollapsed results as a sensitivity check

## The Arellano-Bond Serial Correlation Tests

In the differenced model, $\Delta\epsilon_{it} = \epsilon_{it} - \epsilon_{i,t-1}$

**AR(1) test** — $H_0: \text{Cov}(\Delta\epsilon_{it},\;\Delta\epsilon_{i,t-1}) = 0$

- Even under IID levels errors, $\text{Cov}(\Delta\epsilon_{it},\;\Delta\epsilon_{i,t-1}) = -\sigma^2_\epsilon \neq 0$
- **Should reject** — mechanical from the differencing transformation; failure to reject is the surprise

**AR(2) test** — $H_0: \text{Cov}(\Delta\epsilon_{it},\;\Delta\epsilon_{i,t-2}) = 0$

- Under IID levels errors: $\text{Cov}(\Delta\epsilon_{it},\;\Delta\epsilon_{i,t-2}) = 0$ — **should NOT reject**
- Rejection implies $\text{Cov}(\epsilon_{it},\;\epsilon_{i,t-2}) \neq 0$ in levels — this invalidates $y_{i,t-2}$ as an instrument

**If AR(2) rejects**:

- Use $y_{i,t-3}$ and deeper lags as instruments (sacrifice some efficiency)
- Add further lags of $y$ to the RHS to absorb the serial correlation
- Check AR(3) — how deep does the serial dependence extend?

In R: `mtest(model, order=2)` from `plm`; in Stata: `estat abond`

## One-Step Arellano-Bond: Code and Results

Replicating Arellano and Bond (1991), Table 4(a1):

::: {.panel-tabset}

### R

```{r ab-one-code, eval=FALSE}
library(plm)
data(EmplUK)

Table4.a1 <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) +
             lag(log(wage), 0:1) +
             lag(log(capital), 0:2) +
             lag(log(output), 0:2) |
             lag(log(emp), 2:99),
  data   = EmplUK,
  effect = "twoways",
  model  = "onestep"
)
summary(Table4.a1)
```

### Stata

```stata
use abdata, clear

* One-step difference GMM — Table 4(a1)
xtabond n L(0/1).w L(0/2).(k ys) yr1980-yr1984, ///
    lags(2) vce(robust)

estat abond   /* Arellano-Bond AR tests */
estat sargan  /* Sargan overidentification test */
```

:::

```{r ab-one-run, echo=FALSE}
Table4.a1 <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
             lag(log(capital), 0:2) + lag(log(output), 0:2) |
             lag(log(emp), 2:99),
  data = EmplUK, effect = "twoways", model = "onestep")
summary(Table4.a1)
```

## Two-Step AB with Windmeijer Correction: Code and Results

Replicating Arellano and Bond (1991), Table 4(b):

::: {.panel-tabset}

### R

```{r ab-two-code, eval=FALSE}
Table4.b <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) +
             lag(log(wage), 0:1) +
             log(capital) +
             lag(log(output), 0:1) |
             lag(log(emp), 2:99),
  data   = EmplUK,
  effect = "twoways",
  model  = "twosteps"
)

summary(Table4.b, robust = FALSE)  # Without Windmeijer (matches Stata non-robust)
summary(Table4.b, robust = TRUE)   # With Windmeijer correction — always prefer this
```

### Stata

```stata
* Two-step difference GMM — Table 4(b)
xtabond n L(0/1).w k L(0/1).ys yr1980-yr1984, ///
    lags(2) twostep vce(robust)

estat abond
estat sargan
```

:::

```{r ab-two-run, echo=FALSE}
Table4.b <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
             log(capital) + lag(log(output), 0:1) |
             lag(log(emp), 2:99),
  data = EmplUK, effect = "twoways", model = "twosteps")
summary(Table4.b, robust = TRUE)
```

# System GMM: Arellano-Bover / Blundell-Bond

## The Weak Instruments Problem

Difference GMM works well when $\gamma$ is moderate. Near the unit root, it breaks down.

When $\gamma \approx 1$:

- $y_{it}$ is near-nonstationary; $\Delta y_{i,t-1}$ is nearly white noise
- Lagged levels $y_{i,t-2}$ are very weakly correlated with $\Delta y_{i,t-1}$ — **weak first stage**
- GMM estimates have large bias and unreliable SEs

**Simulation evidence** (Blundell and Bond 1998):

- With $\gamma = 0.9$ and large $\text{Var}(\alpha_i)/\text{Var}(\epsilon_{it})$, the AB estimator is severely biased
- The bias can approach that of the (inconsistent) OLS estimator in magnitude
- The FE/pooled OLS bracket may be very wide, giving little information

**Intuition**: when the process is very persistent, there is little variation in $y_{i,t-1}$ conditional on its own lags — the instrument is weak in exactly the same sense as a weak first stage in 2SLS.

This motivates **System GMM**: augment the difference equation with a levels equation, using lagged differences as additional instruments

## System GMM: Arellano-Bover / Blundell-Bond

**Arellano and Bover (1995)** proposed additional moment conditions from the **levels equation**:

$$y_{it} = \gamma\, y_{i,t-1} + \mathbf{X}_{it}\boldsymbol{\beta} + \alpha_i + \epsilon_{it}$$

Valid instruments for the levels equation: **lagged first differences** $\Delta y_{i,t-1}$

$$\mathbb{E}[\Delta y_{i,t-1}\cdot(\alpha_i + \epsilon_{it})] = 0$$

This holds under **mean stationarity**: $\mathbb{E}[\Delta y_{i2}\cdot\alpha_i] = 0$

i.e., the initial first-difference of $y$ is uncorrelated with the permanent unit effect

**Blundell and Bond (1998)** show this assumption is plausible when $y_{it}$ is stationary and $y_{i0}$ is drawn from the stationary distribution — reasonable for many economic/social processes

System GMM **stacks** the differenced and levels equations, exploiting both sets of conditions jointly:

::: {.nonincremental}
| Equation | Instruments |
|----------|-------------|
| Difference | Lagged levels $y_{i,t-2}, y_{i,t-3},\ldots$ |
| Levels | Lagged differences $\Delta y_{i,t-1}, \Delta y_{i,t-2},\ldots$ |
:::

## Difference GMM vs. System GMM

::: {.nonincremental}
| Feature | Difference GMM (AB) | System GMM (ABB/BB) |
|---------|--------------------|--------------------|
| Equations | Differences only | Differences + Levels |
| DV instruments | Lagged levels | Lagged levels + Lagged diffs |
| Key assumption | IID $\epsilon$; no serial correlation | + Mean stationarity of $y_{i0}$ |
| Best suited for | $\gamma$ small/moderate | $\gamma$ near 1; large $\text{Var}(\alpha_i)$ |
| Risk | Weak instruments if $\gamma \approx 1$ | Invalid if stationarity fails |
| Diagnostic test | AR(2); Sargan/Hansen | AR(2); Sargan/Hansen; diff-in-Hansen |
:::

**The difference-in-Hansen test** (available in `xtabond2`) separately tests the validity of the levels equation instruments — i.e., it tests the mean stationarity assumption

A statistically sensible strategy: estimate both; if estimates are close, prefer System GMM for efficiency. If they differ substantially, investigate why.

## System GMM: Code and Results

Replicating Blundell and Bond (1998), Table 4:

::: {.panel-tabset}

### R

```{r bb-code, eval=FALSE}
Table4.BB <- pgmm(
  log(emp) ~ lag(log(emp), 1) +
             lag(log(wage), 0:1) +
             lag(log(capital), 0:1) |
             lag(log(emp), 2:99) +
             lag(log(wage), 2:99) +
             lag(log(capital), 2:99),
  data           = EmplUK,
  effect         = "twoways",
  model          = "onestep",
  transformation = "ld"   # levels + differences
)
summary(Table4.BB, robust = TRUE)
```

### Stata

```stata
* System GMM via xtdpdsys
xtdpdsys n L(0/1).w L(0/1).k yr1980-yr1984, ///
    lags(1) twostep vce(robust)

estat abond
estat sargan

* Preferred: xtabond2 (Roodman) — adds diff-in-Hansen, collapse, etc.
xtabond2 n L.n L(0/1).w L(0/1).k yr1980-yr1984, ///
    gmm(L.n, lag(2 .) collapse)                  ///
    iv(L(0/1).w L(0/1).k yr1980-yr1984)           ///
    twostep robust small
```

:::

```{r bb-run, echo=FALSE}
Table4.BB <- pgmm(
  log(emp) ~ lag(log(emp), 1) + lag(log(wage), 0:1) +
             lag(log(capital), 0:1) |
             lag(log(emp), 2:99) + lag(log(wage), 2:99) + lag(log(capital), 2:99),
  data = EmplUK, effect = "twoways", model = "onestep", transformation = "ld")
summary(Table4.BB, robust = TRUE)
```

# Specification Testing

## Sargan/Hansen and the Difference-in-Hansen Test

**Sargan test** (one-step, non-robust): $S = N\cdot g_N'\hat{\mathbf{W}}_N g_N \sim \chi^2(L-K)$

**Hansen $J$-test** (two-step, robust): same formula using optimal weighting matrix

::: {.nonincremental}
| $p$-value | Interpretation |
|-----------|----------------|
| $p < 0.05$ | Reject $H_0$ — some instruments likely invalid |
| $0.05$–$0.25$ | Comfortable zone |
| $0.25$–$0.95$ | Acceptable |
| $p > 0.95$ | Too many instruments — test has no power |
:::

**Difference-in-Hansen test**: $\Delta J = J_{system} - J_{difference} \sim \chi^2(L_{extra})$

- Tests the **additional moment conditions** introduced by the levels equation in System GMM
- Rejection suggests the mean stationarity assumption underlying System GMM is violated
- Available in `xtabond2` (Stata); not yet in `plm::pgmm`

Key limitation of all these tests: they are **joint tests** of all included instruments. A rejection does not identify *which* moments fail; a non-rejection does not guarantee all moments are valid.

## The AB AR Tests in Practice

Recall: $\Delta\epsilon_{it} = \epsilon_{it} - \epsilon_{i,t-1}$

**AR(1) in differences is expected**: should reject $H_0$ of no AR(1)

- Under correct specification, adjacent differenced residuals share $-\epsilon_{i,t-1}$
- Failure to reject is the warning sign — check model specification

**AR(2) is the identification-relevant test**: should **not** reject $H_0$

- Rejection implies $\text{Cov}(\epsilon_{it},\epsilon_{i,t-2}) \neq 0$ — violates the identifying assumption that $y_{i,t-2}$ is a valid instrument
- If AR(2) rejects: exclude $t-2$ and instrument from $t-3$ onward

**Also check AR(3)** in long panels to confirm the serial correlation structure

::: {.panel-tabset}

### R

```{r ar-tests, eval=FALSE}
library(plm)
mtest(Table4.b, order = 1)  # AR(1) test
mtest(Table4.b, order = 2)  # AR(2) test — key test
```

### Stata

```stata
* After xtabond / xtdpd / xtabond2
estat abond          /* reports AR(1) and AR(2) by default */
estat abond, lags(3) /* also request AR(3)                 */
```

:::

## A Diagnostic Checklist

::: {.nonincremental}
**Step 1 — Check AR(2)**

If $p < 0.05$: serial correlation in levels. Use instruments dated $t-3$ or later, or extend lag depth.

**Step 2 — Check Sargan/Hansen $J$-test**

If $p < 0.05$: investigate which instrument blocks may be invalid.
If $p > 0.95$: too many instruments — collapse or limit lag depth and retest.

**Step 3 — System GMM only: difference-in-Hansen**

If $p < 0.05$: stationarity assumption suspect — prefer Difference GMM.

**Step 4 — Count instruments vs. groups**

Report $L$ and $N$; confirm $L \leq N$ (Roodman's rule of thumb).

**Step 5 — Plausibility and brackets**

Is $\hat{\gamma} \in (0,1)$? Does FE $< \hat{\gamma}_{AB} <$ OLS? Are signs of $\hat{\boldsymbol{\beta}}$ consistent with theory? Does the long-run effect $\hat{\beta}/(1-\hat{\gamma})$ make substantive sense?
:::

# Bias-Corrected LSDV

## LSDV, Nickell Bias, and the Kiviet/Bruno Correction

**LSDV** (Least Squares Dummy Variables) = within estimator = FE

Nickell's (1981) bias approximation as $N\to\infty$ with $T$ fixed:

$$\text{Bias}(\hat{\gamma}_{LSDV}) \approx -\frac{1+\gamma}{T-1} + O(T^{-2})$$

**Kiviet (1995)** derived higher-order analytic approximations to the **exact finite-sample** bias of LSDV, enabling a bias correction:

$$\hat{\gamma}_{LSDVC} = \hat{\gamma}_{LSDV} - \hat{B}(\hat{\gamma}_{initial})$$

where $\hat{B}$ is estimated using a consistent initial estimate of $\gamma$ from AH, AB, or BB

**Bruno (2005)** extended this to **unbalanced panels**

Bootstrap standard errors are used since analytic SEs for the corrected estimator are complex

**Preferred when**: $N$ is small relative to $T$ (e.g. $N < 50$), where GMM instrument proliferation is especially damaging; or as a robustness check alongside GMM results

## LSDVC in R and Stata

::: {.panel-tabset}

### R

```{r lsdvc-code, eval=FALSE}
# Bias-corrected LSDV is primarily available in Stata
# In R, 'pdynmc' provides related dynamic panel estimators
# install.packages("pdynmc")
library(pdynmc)

# pdynmc estimates linear DPD models with various GMM configurations
# See vignette("pdynmc") for LSDVC-style bias correction options

# As a practical alternative — compare FE to GMM to bracket the true gamma
fe_mod <- plm(
  log(emp) ~ lag(log(emp), 1) + log(wage) + log(capital) + log(output),
  data = EmplUK, model = "within", effect = "twoways"
)
summary(fe_mod)
```

### Stata

```stata
* Bias-corrected LSDV — Bruno (2005)
* initial() options: ah, ab, or bb
* vcov(B) requests B bootstrap replications for SE

xtlsdvc n L.n L(0/1).w L(0/1).k yr1980-yr1984, ///
    initial(ab) vcov(500) lags(1)

* Compare initializations
foreach init in ah ab bb {
    xtlsdvc n L.n L(0/1).w L(0/1).k yr1980-yr1984, ///
        initial(`init') vcov(200) lags(1)
}
```

:::

Estimates are typically between the FE (downward biased) and GMM (larger variance) results — a useful triangulation when $N$ is small.

# Extensions

## Unbalanced Panels and Orthogonal Deviations

**First differences and gaps**: if $y_{i,t-1}$ is missing, $\Delta y_{it}$ is unavailable. Gaps propagate — one missing period can eliminate adjacent observations.

**Orthogonal deviations** (Arellano and Bover 1995): express each observation as a deviation from the mean of all **future available** observations

$$y_{it}^* = c_{it}\left(y_{it} - \frac{1}{T_{it}}\sum_{s>t}y_{is}\right), \quad c_{it} = \sqrt{\frac{T_{it}}{T_{it}+1}}$$

where $T_{it}$ = number of future observations for unit $i$ at time $t$

- Removes $\alpha_i$ (just as differencing does)
- **Does not propagate missing values**: each transformed observation requires only $y_{it}$ and future periods
- Lagged levels remain valid instruments — same logic as for differences
- Typically produces more efficient estimates when the panel has gaps

**Implementation**: `xtabond2, orthogonal` in Stata; orthogonal deviations not available in `plm::pgmm` (use `xtabond2` via `RStata` or `panelvar` for R)

## Panel VAR and Software Notes

**Panel VAR** extends DPD to multivariate systems — all variables treated as jointly endogenous:

$$\mathbf{Y}_{it} = \mathbf{A}_1\mathbf{Y}_{i,t-1} + \cdots + \alpha_i + \boldsymbol{\epsilon}_{it}$$

Applications: macro interdependencies, fiscal–monetary interactions, conflict dynamics

::: {.panel-tabset}

### R

```{r pvar-code, eval=FALSE}
library(panelvar)  # install.packages("panelvar")

pvar_mod <- pvargmm(
  dependent_vars     = c("log_emp", "log_wage"),
  lags               = 1,
  transformation     = "fd",         # first differences
  data               = EmplUK,
  panel_identifier   = c("firm", "year"),
  steps              = "twosteps",
  system_instruments = FALSE,
  max_instr_dependent_vars = 4
)
summary(pvar_mod)
girf(pvar_mod, n.ahead = 10, ma_approx_steps = 10) |> plot()
```

### Stata

```stata
* Panel VAR — Abrigo and Love (2016)
* ssc install pvar

pvar log_emp log_wage, lags(1) fod twostep
pvarirf, oirf mc(500) step(10)
```

:::

::: {.nonincremental}
| R package | Stata command | Coverage |
|-----------|--------------|---------|
| `plm::pgmm` | `xtabond`, `xtdpd` | AB, BB |
| `panelvar` | `xtabond2` | AB, BB, collapse, orthog, diff-in-Hansen |
| `pdynmc` | `xtlsdvc` | LSDVC, extended GMM |
:::

## Slope Heterogeneity: Testing Before You Pool

All AB/BB estimators assume **homogeneous slope vectors** $\boldsymbol{\beta}$ — the same $\gamma$, same wage elasticity, same capital elasticity for every unit. **Pesaran and Smith (1995)** proved that pooling with heterogeneous slopes yields **inconsistent estimates even with no endogeneity**, as $N, T \to \infty$.

**Pesaran-Yamagata (2008) $\tilde{\Delta}$ test** — run $H_0: \boldsymbol{\beta}_i = \boldsymbol{\beta}\;\forall i$ before deciding to pool:

$$\tilde{\Delta} = \sqrt{N}\left(\frac{N^{-1}\hat{S} - K}{\sqrt{2K}}\right) \xrightarrow{d} \mathcal{N}(0,1)$$

where $\hat{S} = \sum_i(\hat{\boldsymbol{\beta}}_i - \hat{\boldsymbol{\beta}}_{FGLS})'\hat{\Sigma}_i^{-1}(\hat{\boldsymbol{\beta}}_i - \hat{\boldsymbol{\beta}}_{FGLS})$

If $H_0$ is rejected, choose an estimator that respects heterogeneity:

::: {.nonincremental}
| Estimator | Description | Use when |
|---|---|---|
| **Mean Group (MG)** (Pesaran-Smith 1995) | Average of $N$ unit regressions | $T$ large; slopes heterogeneous |
| **Pooled Mean Group (PMG)** (Pesaran-Shin-Smith 1999) | LR slopes homogeneous; SR heterogeneous | Cointegrated panel; theory restricts LR |
| **CCE-MG** (Pesaran 2006) | MG + cross-sectional means as common-factor proxies | Cross-sectional dependence present |
| **Dynamic CCE** (`xtdcce2`, Ditzen 2021) | CCE with lagged DV | DPD + cross-sectional dependence |
:::

::: {.panel-tabset}

### R

```{r slope-het, eval=FALSE}
library(plm)

# Pesaran CD test for cross-sectional dependence (prerequisite)
ab_fe <- plm(n ~ lag(n,1) + w + k + ys, data=EmplUK,
             model="within", effect="twoways")
pcdtest(ab_fe)             # Pesaran (2015) CD test

# Mean Group estimator (requires sufficient T per unit)
# pmg() in plm
mg_mod  <- pmg(n ~ lag(n,1) + w + k + ys, data=EmplUK,
               model="mg")   # mean group
pmg_mod <- pmg(n ~ lag(n,1) + w + k + ys, data=EmplUK,
               model="pmg")  # pooled mean group

# Hausman-type test: MG vs PMG (tests LR homogeneity)
phtest(mg_mod, pmg_mod)
```

### Stata

```stata
* Pesaran (2004) CD test
xtcd n             /* cross-sectional dependence test */

* Pesaran-Yamagata slope homogeneity test
xtslope n L.n w k ys   /* ssc install xtslope */

* Mean Group and Pooled Mean Group
xtpmg d.n d.w d.k d.ys, lr(L.n w k ys) ec(ec) full mg
xtpmg d.n d.w d.k d.ys, lr(L.n w k ys) ec(ec) full pmg

* Hausman test: PMG vs MG (tests LR slope homogeneity)
hausman mg_est pmg_est

* Dynamic CCE (handles cross-sectional dependence + lagged DV)
* ssc install xtdcce2
xtdcce2 n L.n w k ys, dcce lr(L.n w k ys)
```

:::

## Extensions to Binary and Count Outcomes

The entire DPD framework so far assumes a **continuous** outcome. But many social science DVs are discrete:

- **Binary**: conflict onset, democratic breakdown, legislation passage, protest occurrence
- **Count**: number of protests, trade deals, policy adoptions, fatalities
- **Ordinal**: Polity/V-Dem democracy scores, treaty compliance levels

**Why linear DPD fails here**:

- Predicted probabilities escape $[0,1]$ (linear probability model)
- The incidental parameters problem (Neyman and Scott 1948): in nonlinear FE models, $\hat{\alpha}_i$ is estimated from $T$ observations and is inconsistent as $T$ fixed, contaminating $\hat{\boldsymbol{\beta}}$
- The **Heckman (1981) initial conditions problem**: the distribution of $y_{i1}$ depends on $\alpha_i$, so we cannot simply condition on $y_{i1}$ as observed

**Wooldridge (2005, JAE)** provides the cleanest solution — **correlated random effects (CRE) conditioning**:

Model $\alpha_i \mid y_{i1},\bar{\mathbf{x}}_i \sim \mathcal{N}(\psi_0 + \psi_1 y_{i1} + \bar{\mathbf{x}}_i\boldsymbol{\psi}_2,\; \sigma^2_a)$, then integrate out $\alpha_i$. This yields a standard **random-effects probit / Poisson** augmented with $y_{i1}$ and unit-means $\bar{\mathbf{x}}_i$ as additional regressors:

$$P(y_{it}=1 \mid y_{i,t-1}, \mathbf{x}_{it}, y_{i1}, \bar{\mathbf{x}}_i) = \Phi\!\left(\gamma y_{i,t-1} + \mathbf{x}_{it}\boldsymbol{\beta} + \psi_1 y_{i1} + \bar{\mathbf{x}}_i\boldsymbol{\psi}_2\right)$$

The initial value $y_{i1}$ proxies the distribution of $\alpha_i$ — no numerical integration beyond what RE probit already does.

::: {.panel-tabset}

### R

```{r dynamic-probit, eval=FALSE}
library(lme4)

# Construct Mundlak/Wooldridge terms
panel_df <- EmplUK |>
  group_by(firm) |>
  mutate(
    n_bin  = as.integer(n > median(n)),  # example binary DV
    n_init = first(n_bin),               # initial value y_{i1}
    w_bar  = mean(w, na.rm = TRUE),      # unit mean of wage
    k_bar  = mean(k, na.rm = TRUE)       # unit mean of capital
  ) |> ungroup()

# Wooldridge (2005) CRE dynamic probit via glmer
mod_cre_probit <- glmer(
  n_bin ~ lag(n_bin) + w + k + n_init + w_bar + k_bar + (1 | firm),
  data = panel_df, family = binomial(link = "probit")
)
summary(mod_cre_probit)

# For counts: CRE Poisson (Blundell-Griffith-Windmeijer 2002)
mod_cre_pois <- glmer(
  emp ~ lag(emp) + w + k + n_init + w_bar + k_bar + (1 | firm),
  data = panel_df, family = poisson
)
```

### Stata

```stata
* Construct Mundlak/Wooldridge terms
bysort id: gen y_init = y_binary[1]
bysort id: egen w_bar  = mean(w)
bysort id: egen k_bar  = mean(k)

* Wooldridge (2005) CRE dynamic probit
xtprobit y_binary L.y_binary w k y_init w_bar k_bar, re

* Average partial effects (critical for interpretation)
margins, dydx(*) predict(pu0)

* For count outcomes: CRE Poisson
xtpoisson emp L.emp w k y_init w_bar k_bar, re
margins, dydx(*) predict(nu0)

* Test: is the initial condition coefficient significant?
* (psi1 != 0 confirms the initial conditions problem is real)
test y_init
```

:::

**Key interpretation note**: coefficients from CRE probit/Poisson are not average partial effects (APEs). Always compute APEs via `margins` (Stata) or the `marginaleffects` package (R) and interpret those — not the latent-index coefficients.

# Applied Example: UK Manufacturing Employment

## The Data: EmplUK / abdata

Based on **Layard and Nickell (1986)**, *Unemployment in Britain*, Economica 53

```
Obs: 1,031     140 UK manufacturing firms, 1976–1984 (unbalanced)

Variable   Description
--------   -----------------------------------------------
id / firm  Firm identifier
year       Year (1976–1984)
emp / n    Employment count / log(employment)
wage / w   Real wage / log(real wage)
capital/k  Gross capital stock / log(capital)
output/ys  Industry output / log(industry output)
yr1980–yr1984   Year dummy indicators
```

::: {.panel-tabset}

### R

```{r data-setup, eval=FALSE}
library(plm)
data(EmplUK)
EmplUK$n  <- log(EmplUK$emp)
EmplUK$w  <- log(EmplUK$wage)
EmplUK$k  <- log(EmplUK$capital)
EmplUK$ys <- log(EmplUK$output)
```

### Stata

```stata
use abdata, clear
* Variables n, w, k, ys already in log form in abdata
xtset id year
xtdescribe       /* inspect panel structure */
xtsum n w k ys   /* within/between/overall variance */
```

:::

**Research question**: How do wages, capital, and industry output shape firm employment? What is the degree of employment persistence, and how quickly does adjustment occur?

## Panel Summary Statistics

```{r panel-summ-show, echo=FALSE}
data(EmplUK)
EmplUK$n  <- log(EmplUK$emp);  EmplUK$w  <- log(EmplUK$wage)
EmplUK$k  <- log(EmplUK$capital); EmplUK$ys <- log(EmplUK$output)
ps2 <- panel.summary(EmplUK[, c("n","w","k","ys")], EmplUK$firm)
ps2$sum.tab |>
  kbl(caption = "Variance decomposition: log employment, wage, capital, output") |>
  kable_styling(font_size = 14)
```

Key observations:

- **Employment** ($n$): roughly 30–40% within — meaningful time-series variation across firms
- **Wage** ($w$) and **Capital** ($k$): dominated by between variation — identification concerns for GMM instruments
- **Output** ($ys$): industry-level variable; nearly all variation is between firms — appropriate as a strictly exogenous regressor

## Specification Choices

**Lag depth for the DV**: AB (1991) uses two lags — test if the second is significant; confirm AR(2) is not rejected with this choice

**Treatment of covariates**:

- $w$ (wage): **predetermined** — firms plausibly adjust wages in response to past employment shocks; instrument with levels at $t-1$ and before
- $k$ (capital): potentially **endogenous** — investment and employment may be jointly determined; instrument with levels at $t-2$ and before
- $ys$ (output): **strictly exogenous** — industry output is not set by an individual firm; enter as $\Delta ys_{it}$

**Time fixed effects**: `effect = "twoways"` in `pgmm`; `yr1980-yr1984` dummies in Stata — controls for macroeconomic shocks common to all firms

**Instrument discipline**: begin with `lag(..., 2:99)`; monitor instrument count and Hansen $p$-value; re-estimate with `2:4` or collapsed as robustness

## Results and Long-Run Interpretation

```{r results-compare, echo=FALSE}
fe <- plm(n ~ lag(n,1) + lag(n,2) + w + lag(w,1) + k + lag(k,1) + lag(k,2) +
              ys + lag(ys,1) + lag(ys,2),
          data = EmplUK, model = "within", effect = "twoways")
ab1 <- Table4.a1;  ab2 <- Table4.b

tab <- data.frame(
  Estimator = c("Fixed Effects (FE)", "AB One-Step", "AB Two-Step (Windmeijer)","System GMM (BB)"),
  `L.emp`   = c(round(coef(fe)[1],3), round(coef(ab1)[1],3), round(coef(ab2)[1],3), round(coef(Table4.BB)[1],3)),
  `L2.emp`  = c(round(coef(fe)[2],3), round(coef(ab1)[2],3), round(coef(ab2)[2],3), "—"),
  `wage`    = c(round(coef(fe)[3],3), round(coef(ab1)[3],3), round(coef(ab2)[3],3), round(coef(Table4.BB)[2],3))
)
tab |> kbl(col.names = c("Estimator","L.emp","L2.emp","Wage (contemp.)"),
           caption = "Selected estimates across estimators") |>
  kable_styling(font_size = 14) |>
  row_spec(0, bold = TRUE)
```

**Key patterns**: FE under-estimates persistence (Nickell bias); System GMM yields the highest $\hat{\gamma}$; AB two-step with Windmeijer SEs is the benchmark for DPD

**Long-run effects from AB two-step (Table 4b)**

In steady state $y^* = \gamma_1 y^* + \gamma_2 y^* + \sum_j \beta_{x,j} x^*$, so:

$$\text{LR effect of } x = \frac{\sum_j \hat{\beta}_{x,j}}{1 - \hat{\gamma}_1 - \hat{\gamma}_2}$$

Sum all lag coefficients on each $x$ (numerator); subtract the **sum** of AR coefficients from 1 (denominator).

```{r lr-effects, echo=TRUE}
# Extract AB two-step coefficients by name for transparency
cf <- coef(Table4.b)

g1  <- cf["lag(log(emp), 1:2)1"];   g2  <- cf["lag(log(emp), 1:2)2"]
bw0 <- cf["lag(log(wage), 0:1)0"];  bw1 <- cf["lag(log(wage), 0:1)1"]
bk  <- cf["log(capital)"]                          # only lag-0 in Table 4b spec
by0 <- cf["lag(log(output), 0:1)0"]; by1 <- cf["lag(log(output), 0:1)1"]

adj_speed <- 1 - g1 - g2          # denominator for all LR effects

cat(sprintf("gamma1 = %6.4f   gamma2 = %6.4f\n", g1, g2))
cat(sprintf("sum(gamma)     = %6.4f\n", g1 + g2))
cat(sprintf("Speed of adj.  = 1 - %6.4f = %6.4f  (~%.0f%% per year)\n",
            g1+g2, adj_speed, 100*adj_speed))
cat(sprintf("\nLR wage  = (bw0 + bw1) / adj  = (%6.4f + %6.4f) / %6.4f = %7.4f\n",
            bw0, bw1, adj_speed, (bw0+bw1)/adj_speed))
cat(sprintf("LR cap   = bk / adj            = %6.4f          / %6.4f = %7.4f\n",
            bk, adj_speed, bk/adj_speed))
cat(sprintf("LR output= (by0 + by1) / adj   = (%6.4f + %6.4f) / %6.4f = %7.4f\n",
            by0, by1, adj_speed, (by0+by1)/adj_speed))
```

# Practical Guidance and A Philosophical Note

## Choosing an Estimator

**Start here — Is there a lagged DV or outcome-to-covariate feedback?**

No → standard FE/RE panel models suffice; stop here

Yes → continue below

**Is $T$ large ($> 20$–$30$)?**

Yes → Nickell bias is small; FE or LSDVC may be sufficient with a robustness check

No → proceed to GMM

**Is $N$ small ($< 50$)?**

Yes → instrument proliferation is severe; use LSDVC, or limit instruments aggressively before applying GMM

**Is $\gamma$ likely near 1 (high persistence)?**

Yes → Difference GMM instruments are weak; prefer System GMM; validate with diff-in-Hansen

No → Difference GMM (AB) is appropriate; report System GMM as robustness

**Always report alongside any GMM result**:

- Instrument count ($L$) vs. number of groups ($N$)
- AR(1) and AR(2) $p$-values
- Sargan/Hansen $p$-value
- FE and pooled OLS for bracketing comparison

## Common Pitfalls

**1. Reporting `lag(y, 2:99)` without checking instrument count**

A Hansen $p$-value of 0.99 is not good news — it signals the test has no power

**2. Treating predetermined covariates as strictly exogenous**

Using $\Delta x_{it}$ as an instrument when $x_{it}$ responds to past $y$ introduces invalid moments

**3. Ignoring the AR(2) test**

If $y_{i,t-2}$ is correlated with the error in levels, it cannot instrument for $\Delta y_{i,t-1}$

**4. Two-step GMM without Windmeijer correction**

Naive two-step standard errors are dramatically under-estimated in short panels; always specify `robust`

**5. Forgetting time fixed effects**

Without year dummies, common macro shocks enter the error and correlate with all lagged instruments

**6. Conflating persistence with long-run elasticity**

$\hat{\gamma}$ measures the speed of adjustment — high $\hat{\gamma}$ means *slow* adjustment, not instability (provided $\hat{\gamma} < 1$)

## Poolers vs. Time-Series Scholars

The core philosophical tension in panel data:

**Poolers** begin with the presumption that data *can* be pooled:

- The composite / population-averaged inference is the target
- Unit effects are nuisances to control for, not the object of inquiry
- Pooling is the default; heterogeneity is corrected, not modelled

**Time-series scholars** begin with deep scepticism of pooling:

- Units are heterogeneous; pooling requires evidence, not assumption
- The unit-specific trajectory is often substantively interesting
- Estimates from short panels with large $N$ describe a fictional average unit

These are **priors and proclivities** — neither is derived from a general rule

DPD/GMM sits in an uncomfortable middle:

- Explicitly motivated by the cross-sectional dimension ($N$ large)
- But identification comes entirely from the time-series dimension of each unit

**The practical upshot**: theory should drive model choice. The choice of estimator is not neutral — it encodes maintained assumptions about the data-generating process. Report multiple estimators, be transparent about what each assumes, and let substantive interpretation guide, not justify, the choice.

## Looking Ahead — GMM-DPD as Causal Claims: The Imai-Kim Connection

GMM-DPD is a consistency-and-identification framework. The causal tradition asks sharper questions: *for whom* and *under what counterfactual* does $\hat{\gamma}$ or $\hat{\boldsymbol{\beta}}$ describe a treatment effect?

**GMM moment conditions are causal claims in disguise**:

- *Strict exogeneity*: $\mathbb{E}[x_{it}\epsilon_{is}]=0\;\forall s,t$ = "outcomes never feed back to $x$" — **no-feedback** in potential-outcomes language
- *Predetermined*: forward feedback forbidden, backward allowed = **sequential ignorability** given observables
- Instrument validity $\mathbb{E}[y_{i,t-2}\Delta\epsilon_{it}]=0$ = an **exclusion restriction**: $y_{i,t-2}$ affects $y_{it}$ only through $\Delta y_{i,t-1}$

**Imai and Kim (2019, AJPS)** show TWFE identifies a causal effect only under two assumptions rarely stated:

::: {.nonincremental}
| Assumption | TWFE | DPD |
|---|---|---|
| **No carryover**: $D_{i,t-s}$ does not directly affect $Y_{it}$ | Maintained silently | Partially relaxed via distributed lags |
| **No reverse feedback**: $Y_{i,t-s}$ does not affect $D_{it}$ | Maintained silently | **Explicitly modelled** — the whole point of the lagged DV |
:::

By including $y_{i,t-1}$ and instrumenting it, DPD moves Assumption 2 *from the error term into the model*. This is the DPD framework's core contribution to causal thinking about panels. Assumption 1 (no carryover of treatment) remains implicitly maintained in most applications and requires distributed lag structures to relax.

## Parallel Trends vs. Sequential Moment Conditions

Both the DiD tradition and the DPD tradition use panel data to recover effects of interest. They represent **two different identification strategies** for related but distinct questions.

::: {.nonincremental}
| | DPD / GMM | DiD / Parallel Trends |
|---|---|---|
| **Question** | How persistent is $y$? What is the effect of $x$ on $y$? | What is the causal effect of a treatment $D$? |
| **Identifying assumption** | Sequential moment conditions: $\mathbb{E}[y_{i,t-s}\Delta\epsilon_{it}]=0$ | Parallel trends: $\mathbb{E}[Y_{it}(0)-Y_{is}(0)]$ equal across groups |
| **Variation used** | Within-unit over time (IV from lagged values) | Across-group difference in trends |
| **Units as** | Exchangeable observations | Treated vs. control groups |
| **Treatment** | $x_{it}$ enters as covariate; may be endogenous | $D_{it}$ is binary or discrete; staggered adoption |
| **Dynamics** | $\hat{\gamma}$ is the object of inference | Often assumed away (no-carryover) |
:::

**The tension**: DiD estimators that ignore $\hat{\gamma}$ are misspecified when outcomes are persistent — the untreated potential outcome $Y_{it}(0)$ itself follows an AR process, and parallel trends in levels does not hold if the pre-treatment trajectories diverge at different rates. Omitting the lagged DV from a DiD specification conflates treatment effects with autoregressive persistence.

## TWFE, ATT(g,t), and the Day 10 Preview

**The TWFE problem**: Goodman-Bacon (2021) shows standard TWFE with staggered adoption is a weighted average of all $2\times 2$ DiD comparisons — including ones where **already-treated units serve as controls for later-treated units**, generating **negative weights** when effects are heterogeneous over time or across cohorts.

**Wooldridge's (2021) unifying result**: TWFE $\equiv$ **two-way Mundlak regression** — the Mundlak (1978) correlated-random-effects device we already know from DPD:

$$y_{it} = \alpha + \tau D_{it} + \underbrace{\bar{y}_i\,\delta + \bar{y}_t\,\phi}_{\text{two-way Mundlak terms}} + u_{it} \;\;\equiv\;\; \text{TWFE}$$

Extending to cohort$\times$time interactions (ETWFE) recovers **Callaway and Sant'Anna's (2021)** causal estimand:

$$\text{ATT}(g,t) = \mathbb{E}\!\left[Y_{it}(g) - Y_{it}(\infty) \;\big|\; G_i = g\right]$$

the average effect for cohort $g$ at time $t$ using never/not-yet-treated controls — heterogeneity-robust by construction.

**The Day 9 → Day 10 handoff**: Omitting $\hat{\gamma}$ from a DiD specification is only valid if $\gamma = 0$ *or* treatment is independent of past outcomes. When the outcome is genuinely persistent, the untreated potential outcome $Y_{it}(0)$ itself follows an AR process and parallel trends in levels need not hold even when it holds in differences. Day 10 addresses when the $\gamma = 0$ restriction can be defended and when a hybrid DPD-DiD approach is required.

::: {.nonincremental}
| Estimator | Estimand | R / Stata |
|---|---|---|
| Callaway-Sant'Anna (2021) | ATT(g,t) | `did`; `csdid` |
| de Chaisemartin-D'Haultfoeuille (2020) | ATT for switchers | `DIDmultiplegt`; `did_multiplegt` |
| Sun-Abraham (2021) | CATT(g,t) | `fixest::sunab()`; `eventstudyinteract` |
| Wooldridge (2021, 2023) | ATT aggregates (ETWFE) | `etwfe`; `jwdid` |
:::

## References

::: {.nonincremental .smaller}
**Core DPD and GMM**

- Anderson, T.W. and Hsiao, C. (1981). Estimation of dynamic models with error components. *JASA*, 76, 598–606.
- Arellano, M. and Bond, S. (1991). Some tests of specification for panel data. *Review of Economic Studies*, 58, 277–297.
- Arellano, M. and Bover, O. (1995). Another look at the instrumental variable estimation of error-components models. *Journal of Econometrics*, 68, 29–51.
- Blundell, R. and Bond, S. (1998). Initial conditions and moment restrictions in dynamic panel data models. *Journal of Econometrics*, 87, 115–143.
- Bruno, G.S.F. (2005). Approximating the bias of the LSDV estimator for dynamic unbalanced panel data models. *Economics Letters*, 87, 361–366.
- Holtz-Eakin, D., Newey, W. and Rosen, H.S. (1988). Estimating vector autoregressions with panel data. *Econometrica*, 56, 1371–1395.
- Nickell, S. (1981). Biases in dynamic models with fixed effects. *Econometrica*, 49, 1417–1426.
- Roodman, D. (2009). How to do xtabond2. *Stata Journal*, 9, 86–136.
- Wawro, G. (2002). Estimating dynamic panel data models in political science. *Political Analysis*, 10, 25–48.
- Windmeijer, F. (2005). A finite sample correction for the variance of linear efficient two-step GMM estimators. *Journal of Econometrics*, 126, 25–51.

**Slope Heterogeneity and Cross-Sectional Dependence**

- Ditzen, J. (2021). Estimating long-run effects and the exponent of cross-sectional dependence: an update to `xtdcce2`. *Stata Journal*, 21, 687–707.
- Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica*, 46, 69–85.
- Pesaran, M.H. (2006). Estimation and inference in large heterogeneous panels with a multifactor error structure. *Econometrica*, 74, 967–1012.
- Pesaran, M.H., Shin, Y. and Smith, R.P. (1999). Pooled mean group estimation of dynamic heterogeneous panels. *JASA*, 94, 621–634.
- Pesaran, M.H. and Smith, R.P. (1995). Estimating long-run relationships from dynamic heterogeneous panels. *Journal of Econometrics*, 68, 79–113.
- Pesaran, M.H. and Yamagata, T. (2008). Testing slope homogeneity in large panels. *Journal of Econometrics*, 142, 50–93.

**Nonlinear and Limited Dependent Variable DPD**

- Blundell, R., Griffith, R. and Windmeijer, F. (2002). Individual effects and dynamics in count data models. *Journal of Econometrics*, 108, 113–131.
- Heckman, J.J. (1981). The incidental parameters problem and the problem of initial conditions. In Manski and McFadden (eds.), *Structural Analysis of Discrete Data*. MIT Press.
- Wooldridge, J.M. (2005). Simple solutions to the initial conditions problem in dynamic, nonlinear panel data models with unobserved heterogeneity. *Journal of Applied Econometrics*, 20, 39–54.
- Wooldridge, J.M. (2023). Simple approaches to nonlinear difference-in-differences with panel data. *Econometrics Journal*, 26, C31–C66.

**Causal Identification and DiD (Day 10 preview)**

- Callaway, B. and Sant'Anna, P.H.C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics*, 225, 200–230.
- de Chaisemartin, C. and D'Haultfoeuille, X. (2020). Two-way fixed effects estimators with heterogeneous treatment effects. *American Economic Review*, 110, 2964–2996.
- Goodman-Bacon, A. (2021). Difference-in-differences with variation in treatment timing. *Journal of Econometrics*, 225, 254–277.
- Imai, K. and Kim, I.S. (2019). When should we use unit fixed effects regression models for causal inference with longitudinal data? *American Journal of Political Science*, 63, 467–490.
- Sun, L. and Abraham, S. (2021). Estimating dynamic treatment effects in event studies with heterogeneous treatment effects. *Journal of Econometrics*, 225, 175–199.
- Wooldridge, J.M. (2021/2025). Two-way fixed effects, the two-way Mundlak regression, and difference-in-differences estimators. *Empirical Economics*. SSRN 3906345.
:::
